---
title: "Are causal discovery algorithms useful? How can we tell?"
subtitle: "aligning evaluation metrics with research goals" 
author: "Nicholas Putney"
editor: visual
bibliography: WageningenPhDReadingList.json
---
# Introduction
## Causal graphs can be useful!
A causal graph provides a compact representation of the causal relationships of the system being studied and allows one to reason about what would happen under changes to the system. 

For example, take the two graphs below. Graph $A$ suggests that increasing coffee consumption will indeed increase lifespan, via increasing antioxidant activity. Graph $B$, however, suggests that changing coffee intake would have no impact on lifespan and the only reasons why we see a relationship between the two is due to their relationship with genetics. For example, certain genes may "cause" people to drink more coffee and be related to longevity. 

```{r global_options, include=FALSE}
# This is a way to set options for all code chunks at once
# Note that you can also dynamically control options by setting them to a value
# DPI setting increased for Word output, cairo is anti aliasing
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      dev="png", fig.width = 8, fig.height = 4, dpi=200, dev.args=list(type="cairo"))
```


```{r}
library(caugi)
library(patchwork)
```


```{r}
# Create a simple DAG
g1 <- caugi(
  Coffee %-->% Antiox. %-->% Lifespan,
  Genetics %-->% Lifespan,
  class = "DAG"
)

g2 <- caugi(
  Genetics %-->% Coffee + Lifespan,
  Antiox. %-->% Lifespan,
  class = "DAG"
)
```

```{r}
node_style = list(
    fill = "#30c9f0", # Fill color
    col = "black", # Border color
    lwd = 2, # Border width
    padding = 3, # Text padding (mm)
    size = 1, # Size multiplier
    by_node = list(Coffee = list(fill = "red"), Lifespan = list(fill = "#a4f28a"))
)

edge_style = list(
    col = "black", # Edge color
    lwd = 3, # Edge width
    arrow_size = 4 # Arrow size (mm)
)

label_style = list(
    col = "black", # Text color
    fontsize = 12, # Font size
    fontface = "bold", # Font face
    fontfamily = "sans" # Font family
)

tier_style = list(
    fill = "white",
    col = "gray50",
    lty = 0,
    alpha = 0.3
)
```


```{r, fig.width=10, fig.height=6}
tiers_g1 <- c(Coffee = 1, Antiox. = 2, Lifespan = 3, Genetics = 3)


p1 <- plot(g1,
     layout = "sugiyama",
     #tiers = tiers_g1,
     #tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style,
  main = "A")
p2 <- plot(g2,
  layout = "sugiyama",
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style,
  main = "B")

p1 + p2
```
Beyond being a convenient representation, causal graphs can also be used as a formal tool to guide (observational) data analysis. For example, if graph $A$ is your chosen representation, then estimating the causal effect of coffee on lifespan only requires collecting data on coffee consumption and lifespan, because there are no confounders. On the other hand, graph $B$ would require one to "adjust" somehow for genetics by including it in the statistical model. Note that these graphs provide information about "identification" of causal effects and not how to "estimate" them. For more info about using causal graph in applied research, see e.g. @tennantUseDirectedAcyclic2021. 

## Where do the causal graphs come from?
The work of @tennantUseDirectedAcyclic2021 also highlights an important challenge: how does one construct the causal graph in the first place? Typically it is done via consultation with substantive experts and by searching literature. However, different experts can produce rather different graphs and evidence to construct such a detailed graph may be lacking altogether in some fields. 

For example, due to high degrees of confounding and inherent difficulties with conducting randomized controlled trials, causal interpretation in nutrition research is hard. There are also challenges in the field of healthy longevity, where the goal is to determine which exposures or treatments will contribute to maintaining a high quality of life into old age. 

## Causal discovery to the rescue?
Does this mean that estimating causal effects from observational data is not possible for these fields? Not necessarily - different causal structures give rise to different testable patterns in the data. One idea then is to first learn the causal graph from the data and then use it to estimate the causal effect of interest. In fact, methods for learning (some parts of) a causal graph from observational data have been developed since the 1980s. This task of inferring a causal graph from data is generally referred to as "causal discovery" or "structure learning". A relatively gentle introduction to the field of causal discovery was written by @vowelsDyaDAGsSurvey2022. 

## Maybe not...
While these methods have a long history and new methods are being developed at an increasing pace, their adoption in applied research has been slow. Part of the reason may be their disconnection from the literature on causal effect estimation. The disconnection may be partly due to historically reasons, as causal effect estimation methods developed in the fields of biostatistics, epidemiology, economics, and sociology, while causal discovery methods have their roots in computer science, AI, and mathematics. Furthermore, the goals are often different: the vast majority of causal discovery methods infer a complete graph structure, whereas the aim in applied research is more targeted (e.g. the effect of a single exposure on an outcome).

Furthermore, most causal discovery methods can recover the "Markov equivalence class" of a graph, which means the output may contain undirected edges. Given that a single edge direction can have meaningful change on causal conclusions, this may limit its usefulness in guiding data analyses in applied research. 

## Objective of this post
The goal of this post is the following: to explore how "useful" causal discovery methods can be in applied research. Useful is defined in the context of below "workflow":

1. Formulate a research question
2. Translate research question into a mathematical object (i.e. causal estimand)
3. Write down assumptions about the system in the form of a causal graph
4. From the graph, determine which information is necessary to answer the research question
5. Choose an appropriate statistical model, run the model, conduct sensitivity analyses, interpret results, etc.

Causal discovery has strong potential to aid in 3., especially when the system under study is not well understood. 

# Simulating data
The data generating processes considered here will be kept simple to see how discovery methods perform in the best case scenario.

```{r}
#| include: false
# Libraries
library(dplyr)
library(pcalg)
library(causalDisco)
library(SID)
```

I generated multivariate normal data, with only five variables and a probability of connecting an edge set to 0.5. Again, the idea is to keep things simple. 
```{r, echo=TRUE}
# Generate a random ground truth DAG 
n <- 10000
e_prob <- 0.5
p <- 10

set.seed(2)
gt_sim <- generate_graph(n = p, p = e_prob, class = "DAG")
df <- simulate_data(gt_sim, n, coef_range = c(-1,1))
```


# Running causal discovery algorithms
I now run the PC algorithm using the Fisher-Z test (see e.g. https://www.statisticshowto.com/fisher-z/).
```{r, echo = FALSE}
# Estimate (CP)DAG from simulated data
pc_pcalg <- causalDisco::pc(engine = "pcalg", test = "fisher_z", alpha = 0.05)
pc_result <- disco(data = df, method = pc_pcalg)
est <- pc_result$caugi
```

```{r}
# try some more algorithms...
```


# Evaluating algorithms

## Metrics
We now need some way of comparing the inferred graph to the true graph. Below is a brief background and motivation for some of the commonly used metrics. It is perhaps a bit long but I think aligning metrics with the research question is important, so it is worth the time. 

### Structural hamming distance
The most common metric used is the structural hamming distance (SHD), which counts the number of edge additions, deletions, or reversals necessary to go from one graph to the other. It is fast to calculate, fairly intuitive, and can be used to compare both DAGs and PDAGs (which can contain undirected edges).

{{< dstart summary = "SHD formula." >}}

Define $\mathcal{G}$ as the true graph, $\mathcal{H}$ as the estimated graph, and $(i,j) \in V \times V$ as edges.

The SHD can be defined as 

$$\text{SHD}(\mathcal{G}, \mathcal{H}) = \sum_{i,j \in V \times V} \mathbb{1}(\text{EdgeType}_{\mathcal{G}}(i,j) \ne \text{EdgeType}_{\mathcal{H}}(i,j)),$$

where $\mathbb{1}$ is the indicator function.    
{{< dstop >}}

Remember that the context of this post is determining how useful causal discovery algorithms can be for helping practitioners answer specific causal questions. The SHD does not include any notion of causality in its definition. 

```{r}
tiers_g1 <- c(Coffee = 1, Antiox. = 2, Lifespan = 3, Genetics = 3)


p1 <- plot(g1,
     layout = "sugiyama",
     #tiers = tiers_g1,
     #tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style,
  main = "A")
p2 <- plot(g2,
  layout = "sugiyama",
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style,
  main = "B")
```


```{r}
node_style = list(
    fill = "#30c9f0", # Fill color
    col = "black", # Border color
    lwd = 2, # Border width
    padding = 3, # Text padding (mm)
    size = 1, # Size multiplier
    by_node = list(X1 = list(fill = "red"), X3 = list(fill = "#a4f28a"))
)

tiers_g2 <- c(X1 = 1, X2 = 2, X4 = 2, X3 = 3)
```


```{r, fig.width=12, fig.height = 4}
gt <- caugi(
  X4 %-->% X1 %-->% X2 %-->% X3,
  X4 %-->% X3
)

h1 <- caugi(
  X4 %-->% X1 %-->% X2 %-->% X3,
  X4 %-->% X3 + X2
)

h2 <- caugi(
  X1 %-->% X2 %-->% X3,
  X1 %-->% X4,
  X4 %-->% X3 
)

p_gt <- plot(gt,
             layout = "tiered",
             tiers = tiers_g2,
             tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style, main = "G")

p_h1 <- plot(h1,
             layout = "tiered",
             tiers = tiers_g2,
             tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style, main = "H1")

p_h2 <- plot(h2,
             layout = "tiered",
             tiers = tiers_g2,
             tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style, main = "H2")
```

Take the below plots as an example of how this could be problematic. $\mathcal{G}$ is the true graph and $\mathcal{H_1}$ and $\mathcal{H_2}$ are graphs each one change away from $G$ (i.e. $\text{SHD}(\mathcal{G}, \mathcal{H_1}) = \text{SHD}(\mathcal{G}, \mathcal{H_2}) = 1$). 

```{r, fig.width=12, fig.height = 4}
p_gt + p_h1 + p_h2
```

Even though the SHD for $\mathcal{H_1}$ and $\mathcal{H_2}$ are the same, the graphs lead to different causal conclusions. Let's say we are interested in estimating the effect of $X_1$ on $X_3$, $P(X_3 | \text{do}(X_3 = c))$. According to $\mathcal{G}$, a correct adjustment set (think: should include in the statistical model) would be $\{X_4\}$. The extra arrow in $\mathcal{H_1}$ does not change this adjustment set. On the other hand, $\mathcal{H_2}$  suggests that we should not adjust for $X_4$, which would lead to an incorrect estimate. 

Despite this drawback, the SHD is reported in most papers that evaluates causal discovery algorithms. 

{{< dstart summary = "Other metrics" >}}
Other metrics that simply compare edge differences are also commonly reported, such as precision, recall, false discovery rate, etc.
{{< dstop >}}

### Structural intervention distance
Using an example similar to the one above as motivation, @petersStructuralInterventionDistance2015 propose the "structural intervention distance" (SID). The SID involves testing if the adjustment sets implied by some estimated graph $\mathcal{H}$ are valid in the true graph $\mathcal{G}$. More specifically, the authors propose using the parents of the variable being intervened upon as the adjustment set to compare between graphs. Then, for for every ($i,j$), it is tested if the parents of $X_i$ in $\mathcal{H}$, $\text{PA}^\mathcal{H}_{X_i}$, are a valid adjustment for calculating $p^\mathcal{G}(X_j | \text{do}(X_i))$. If $\text{PA}^\mathcal{H}_{X_i}$ is not valid, then the SID increases by 1. 


{{< dstart summary = "SID formula" >}}

$$
\text{SID}(\mathcal{G}, \mathcal{H}) = \#\{(i,j), i \ne j : \text{PA}_i^\mathcal{H} \notin \mathcal{A}^\mathcal{G}(i,j)\},

$$
where $\mathcal{A}^\mathcal{G}(i,j)$ are the collection of valid adjustment sets in the true DAG $\mathcal{G}$ for estimating the total effect of $X_i$ on $X_j$.

{{< dstop >}}

I think this evaluation metric provides a better sense of how "useful" a causal discovery algorithm may be, given that it explicitly uses causality in its definition. It should be said, though, that the parents of $X_i$ are not the only way to define an adjustment set. It's also important to recognize that this metric is "global", as each $(i,j)$ contributes equally. This is in contrast to the motivating context, where the main interest is in the effects of a limited number of exposures on a limited number of outcomes. 

Below is a simple example below where an edge reversal later in the topological order leads to a high SID. The only difference between $\mathcal{G}$ and $\mathcal{H}$ is the change from $X_5 \rightarrow X_6$ to $X_5 \leftarrow X_6$. 

```{r}
tiers_g3 <- c(X1 = 1, X2 = 2, X7 = 2, X3 = 3, X4 = 4, X5 = 5, X6 = 6)

gt <- caugi(
  X1 %-->% X2 %-->% X3 %-->% X4 %-->% X5 %-->% X6,
  X7 %-->% X1 + X3,
  class = "DAG"
)

h1 <- caugi(
  X1 %-->% X2 %-->% X3 %-->% X4 %-->% X5,
  X7 %-->% X1 + X3,
  X6 %-->% X5,
  class = "DAG"
)

p_gt <- plot(gt,
             layout = "tiered",
             tiers = tiers_g3,
             tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style, main = "G")

p_h <- plot(h1,
             layout = "tiered",
             tiers = tiers_g3,
             tier_style = tier_style,
  node_style = node_style,
  edge_style = edge_style,
  label_style = label_style, main = "H")

p_gt / p_h
```
```{r}
#| eval: false
#| include: false
# Using type = parent and normalized = FALSE is equivalent to the SID
aid(gt, h1, type = "parent", normalized = FALSE)

# Verify that above gives same result as SID
# gt_adj <- as_adjacency(gt)
# h1_adj <- as_adjacency(h1)[rownames(gt_adj),]
# h1_adj <- h1_adj[,colnames(gt_adj)]
# 
# gt_adj
# h1_adj
# 
# structIntervDist(gt_adj, h1_adj)
```

Comparing these two graphs, $\text{SID}(\mathcal{G}, \mathcal{H}) = 7$. Despite this, $\mathcal{H}$ gives a valid adjustment set for estimating the total effect of $X_1$ on $X_3$.

```{r}
# Using type = parent and normalized = FALSE is equivalent to the SID
aid(gt, h1, type = "parent", normalized = FALSE)
```

### Adjustment identification distance
A similar metric is the adjustment identification distance (AID) which includes the SID as a special case. 

### Comparing causal effects
The SID and AID are still not exactly aligned with the context laid out in introduction. What we really want to know is: if we use the adjustment sets implied by our estimated graph, how off will our causal effect estimate be (and with what degree of certainty). 

We can test this! The idea will be to calculate the average bias over each possible total causal effect using the estimated graph. The first step is calculating the matrix of causal coefficients, $B$, in $X = BX + \epsilon$. Afterwards, the matrix of all total causal effects for each $(i,j), i \ne j$ is given by $T = (I - B)^{-1} - I$. The entries $i,j$ in $T^{'}$ give $\mathbb{E}[X_j | \text{do}(X_i = 1)] - \mathbb{E}[X_j | \text{do}(X_i = 0)]$. 

Now that we have some idea of how the algorithms will be evaluated, let's move on to the actual evaluation part.

## Evaluation
```{r}
# Some helper functions...
estimate_B <- function(dag, data) {
  vars <- dag@nodes$name
  p <- length(vars)

  # Ensure ordering consistency
  data <- data[, vars]
  S <- cov(data)
  B <- matrix(
    0,
    nrow = p,
    ncol = p,
    dimnames = list(vars, vars)
  )

  for(child in vars) {
    pa <- parents(dag, child)
    if(length(pa) == 0)
      next

    # Parent covariance matrix
    S_pp <- S[pa, pa, drop = FALSE]

    # Parent-child covariance vector
    S_py <- S[pa, child, drop = FALSE]

    # Regression coefficients
    beta <- solve(S_pp, S_py)

    # Store 
    B[child, pa] <- beta
  }

  return(B)
}
```


```{r, echo=TRUE}
# Generate a random ground truth DAG 
n <- 10000
e_prob <- 0.3
p <- 10

set.seed(106)
gt_sim <- generate_graph(n = p, p = e_prob, class = "DAG")
df <- simulate_data(gt_sim, n, coef_range = c(-1,1))
```


```{r, echo=TRUE}
# Generate a random ground truth DAG 
n <- 10000
e_prob <- 0.5
p <- 10
#set.seed(106)
```


```{r, echo=TRUE}
#set.seed(106)
gt_sim <- generate_graph(n = p, p = e_prob, class = "DAG")
df <- simulate_data(gt_sim, n, coef_range = c(0.3, 1))

pc_pcalg <- causalDisco::pc(engine = "pcalg", test = "fisher_z", alpha = 0.05, u2pd = "rand")
pc_result <- disco(data = df, method = pc_pcalg)
est <- pc_result$caugi

# Selecting one DAG from estimated Markov equivalence class
# test <- pcalg::pdag2allDags(as_adjacency(est))
# nrow(test$dags)
```


```{r, echo=TRUE}
est_dag <- dag_from_pdag(est)
d_dag <- calc_distance_metrics(gt = gt_sim, est = est_dag, graph_only = FALSE, df = df, PDAG = FALSE)
d_dag


```

# Running causal discovery algorithms
I now run the PC algorithm using the Fisher-Z test (see e.g. https://www.statisticshowto.com/fisher-z/).
```{r, echo = FALSE}
# Estimate (CP)DAG from simulated data
pc_pcalg <- causalDisco::pc(engine = "pcalg", test = "fisher_z", alpha = 0.05, u2pd = "retry")
ges_pcalg <- causalDisco::ges(engine = "pcalg", score = "sem_bic")

pc_result <- disco(data = df, method = pc_pcalg)
ges_result <- disco(data = df, method = ges_pcalg)

est_pc <- pc_result$caugi
est_ges <- ges_result$caugi

# Selecting one DAG from estimated Markov equivalence class
est_dag <- dag_from_pdag(est_pc)
est_dag <- dag_from_pdag_2(est_pc)
d_dag <- calc_distance_metrics(gt = gt_sim, est = est_dag, graph_only = FALSE, df = df, PDAG = FALSE)
d_dag
```

```{r}
offdiag <- function(M){
  M[row(M) != col(M)]
}

sign_tol <- function(x, tol = 1e-8){
  ifelse(abs(x) < tol, 0, sign(x))
}

sign_accuracy <- function(T_true, T_est, tol = 1e-8){

  true <- offdiag(T_true)
  est  <- offdiag(T_est)

  mean(
    sign_tol(true, tol) ==
    sign_tol(est, tol)
  )
}

nonzero_effects_recall <- function(
    T_true,
    T_est,
    tol = 1e-8
){

  # Exclude diagonals
  idx_offdiag <- row(T_true) != col(T_true)

  true <- T_true[idx_offdiag]
  est  <- T_est[idx_offdiag]

  # True nonzero effects
  idx_true_nonzero <- abs(true) >= tol

  # Was estimated as nonzero?
  recovered <- abs(est[idx_true_nonzero]) >= tol 

  mean(recovered)
}

true_effects_recall <- function(
    T_true,
    T_est,
    rel_threshold = 0.2,
    abs_threshold = NULL,
    tol = 1e-8
){

  # Remove diagonals
  idx_offdiag <- row(T_true) != col(T_true)

  true <- T_true[idx_offdiag]
  est  <- T_est[idx_offdiag]

  # Keep only true nonzero effects
  idx_true_nonzero <- abs(true) >= tol

  true <- true[idx_true_nonzero]
  est  <- est[idx_true_nonzero]

  # Relative error
  if(!is.null(rel_threshold)){

    rel_err <- abs(est - true) / abs(true)

    return(
      mean(rel_err <= rel_threshold)
    )
  }

  # Absolute error
  if(!is.null(abs_threshold)){

    abs_err <- abs(est - true)

    return(
      mean(abs_err <= abs_threshold)
    )
  }

  stop("Provide rel_threshold or abs_threshold")
}

mean_absolute_bias <- function(
    T_true,
    T_est,
    tol = 1e-8
){

  true <- offdiag(T_true)
  est  <- offdiag(T_est)

  idx <- abs(true) >= tol

  mean(abs(est[idx] - true[idx]))
}
```



```{r}
compute_total_effect_matrix <- function(B){
  p <- nrow(B)
  t(solve(diag(p) - B) - diag(p))
}

get_all_dags <- function(cg){
  cg_adj <- as_adjacency(cg)
  all_dags <- pdag2allDags(cg_adj, verbose = FALSE)
  p <- nrow(cg_adj)
  lapply(seq_len(nrow(all_dags$dags)), function(i){
    dag_adj <- matrix(
      all_dags$dags[i, ],
      p, p,
      byrow = TRUE,
      dimnames = list(rownames(cg_adj), colnames(cg_adj))
    )
    as_caugi(dag_adj)
  })
}

# gt and est should be caugi
compute_graph_metrics <- function(gt, est){
  list(
    shd = shd(gt, est),
    sid = aid(gt, est, type = "parent", normalize = FALSE),
    aid = aid(gt, est, type = "ancestor", normalize = FALSE),
    aid_norm = aid(gt, est, type = "ancestor", normalize = TRUE)
  )
}

# Total effect matrices
calculate_total_effect_matrices <- function(cg, df, PDAG = TRUE){
  if(PDAG){
    all_dags <- get_all_dags(cg)
    T_ls <- lapply(all_dags, function(dag){
      B_cg <- estimate_B(dag = dag, data = df)
      T_cg <- compute_total_effect_matrix(B_cg)
      unlist(T_cg)
    })
    return(T_ls)
  } else {
    B_cg <- estimate_B(dag = cg, data = df)
    T_cg <- compute_total_effect_matrix(B_cg)
    return(T_cg)
  }
}

gt <- gt_sim

# Total effect distance
calculate_total_effect_distance <- function(gt, est, df, PDAG = TRUE, norm = c("cor", "mae", "")){

  # True total effect matrix
  #gt_nel <- igraph::as_graphnel(caugi::as_igraph(gt))
  #B_true <- graphNEL_to_B(gt_nel)
  #T_true <- compute_total_effect_matrix(B_true)
  T_true <- calculate_total_effect_matrices(cg = gt, df = df, PDAG = FALSE)

  # Estimated total effect matrices
  T_est <- calculate_total_effect_matrices(cg = est, df = df, PDAG = PDAG)
  
  compare_effects <- function(T_hat){

    idx <- row(T_true) != col(T_true)

    true <- T_true[idx]
    est  <- T_hat[idx]
    tol = 0.01
    metrics <- list(
      nz_recall = nonzero_effects_recall(T_true, T_est, tol = tol),
      true_effect_recall = true_effects_recall(T_true, T_est, abs_threshold = 0.01, rel_threshold = NULL, tol = tol),
      avg_bias = mean_absolute_bias(T_true, T_est, tol = tol)
      # mae = mean(abs(true - est)),
      # rmse = sqrt(mean((true - est)^2)),
      # cor = cor(true, est),
      # sign_acc = mean(sign(true) == sign(est))
    )
    return(metrics)
  }
  
  # Compare
  if(PDAG){
    lapply(T_est, compare_effects)
  } else {
    compare_effects(T_est)
  }
}
```


```{r}
# Main metrics function
calc_distance_metrics <- function(gt, est, graph_only = FALSE, df = NULL, PDAG = TRUE){
  if(PDAG){
    all_dags <- get_all_dags(est)
    metrics <- lapply(all_dags, function(dag){
      metrics_i <- compute_graph_metrics(gt, dag)
      if(!graph_only){metrics_i$tce <- calculate_total_effect_distance(gt = gt, est = dag, df = df, PDAG = FALSE)}
      metrics_i
    })
  } else {
    metrics <- compute_graph_metrics(gt, est)
    if(!graph_only){
      metrics_tce <- calculate_total_effect_distance(gt, est, df, PDAG = FALSE)
      metrics <- c(metrics, metrics_tce)
      }
  }
  return(metrics)
}
```

```{r}
# Estimating multiset of distances
#d_pdag <- calc_distance_metrics(gt = as_caugi(gt_sim), est = est, graph_only = FALSE, df = df, PDAG = #TRUE)

# Selecting one DAG from estimated Markov equivalence class
est_dag <- dag_from_pdag_2(est)
est_dag <- dag_from_pdag(est)
d_dag <- calc_distance_metrics(gt = gt_sim, est = est_dag, graph_only = FALSE, df = df, PDAG = FALSE)
d_dag
```

## Comparison to randomly generated DAGs
It is a bit difficult to understand how well the algorithms actually performed. One way to evaluate this is to compare the algorithms to a graph where edges are chosen randomly. I will randomly generate a DAG and compute its distance from the ground truth a number of times to create a sort of null distribution with which the sample metrics can be compared.

The random DAGs will be parameterized with an edge probability equal to that of the true DAG. Because of this these random DAGs are "too good" as the degree of sparsity will never be known in reality. Later, we will repeat this process with a different sparsities to see how the results change.

```{r}
# Here the end result should be the metrics for each algorithm. Then we go to the null distributions
```

```{r}
# Function for generating random DAGs
generate_random_dag <- function(p, e_prob){
  #V_names <- as.character(sample(1:p)) # this determines topological order
  #g_rand <- pcalg::randomDAG(p, prob = e_prob, lB=0.1, uB=1, V = V_names)
  g_rand <- generate_graph(n = p, p = e_prob, class = "DAG")
  return(g_rand)
}

# Function for calculating distance between ground truth and random DAG
## gt must be a caugi graph
calc_distance_random_dag <- function(gt, e_prob, df){
  p <- length(gt)
  g_rand <- generate_random_dag(p = p, e_prob = e_prob)
  metrics <- calc_distance_metrics(gt, g_rand, graph_only = FALSE, df = df, PDAG = FALSE)
  return(metrics)
}
```

```{r}
library(purrr)
n <- 300
rand_eval_results <- map_dfr(seq_len(n), ~ calc_distance_random_dag(gt = gt_sim, e_prob = 0.5, df = df))
#rand_eval_results
cor(rand_eval_results)
```


```{r, fig.width=12, fig.height=8}
library(ggplot2)
library(tidyr)
library(ggpubr)
rand_eval_results_long <- rand_eval_results %>%
  pivot_longer(cols = everything(),
               names_to = "metric",
               values_to = "value")

quantile_lines <- rand_eval_results_long %>%
  dplyr::group_by(metric) %>%
  dplyr::summarise(
    q025 = quantile(value, probs = 0.025, na.rm = TRUE),
    q975 = quantile(value, probs = 0.975, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c(q025, q975),
    names_to = "line_type",
    values_to = "xintercept"
  )

d_dag_lines <- tibble::tibble(
  metric = names(d_dag),
  line_type = "d_dag",
  xintercept = as.numeric(d_dag)
)

vline_data <- dplyr::bind_rows(quantile_lines, d_dag_lines)
```


```{r, fig.width=12, fig.height=8}
rand_eval_results_long %>% ggplot(aes(x = value)) +
  geom_histogram(bins = 30, alpha = 0.5) +
  geom_vline(
    data = vline_data,
    aes(xintercept = xintercept, color = line_type),
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  scale_color_manual(values = c(
    q025 = "blue",
    q975 = "blue",
    d_dag = "red"
  )) +
  
  scale_linetype_manual(values = c(
    q025 = "dashed",
    q975 = "dashed",
    d_dag = "solid"
  )) +
  
  scale_linewidth_manual(values = c(
    q025 = 0.7,
    q975 = 0.7,
    d_dag = 1.5
  )) +
  facet_wrap(~ metric, scales = "free") +
  theme_bw()


# hist(rand_eval_results$d_tce)
# hist(rand_eval_results$aid)
# hist(rand_eval_results$sid)
# hist(rand_eval_results$aid_norm)



```


## But did I get lucky?
Now I repeated this with 20 different randomly chosen "ground truth DAGs"
```{r}
# Repeat the same workflow over many simulated ground-truth DAGs (all with p = 10)

method = method_specs[[3]]$method
evaluate_one_ground_truth <- function(gt_id, method, algorithm){
  attempt <- 1L

  repeat {
    if (attempt > max_attempts) {
      stop(
        paste0(
          "Could not obtain a DAG extension for algorithm = ",
          algorithm,
          ", gt_id = ",
          gt_id,
          " after ",
          max_attempts,
          " attempts."
        )
      )
    }

    result_i <- tryCatch({
      gt_i <- generate_graph(n = p_gt, p = e_prob_gt, class = "DAG")
      df_i <- simulate_data(gt_i, n_obs, coef_range = c(-1, 1))

      est_i <- disco(data = df_i, method = method)$caugi
      est_dag_i <- if (caugi::is_dag(est_i) & est_i@graph_class == "PDAG") mutate_caugi(est_i, "DAG") else dag_from_pdag_2(est_i)
      d_dag_i <- calc_distance_metrics(gt = gt_i, est = est_dag_i, graph_only = FALSE, df = df_i, PDAG = FALSE)

      rand_i <- map_dfr(
        seq_len(n_null),
        ~ calc_distance_random_dag(gt = gt_i, e_prob = e_prob_null, df = df_i)
      )

      rand_i_long <- rand_i %>%
        tidyr::pivot_longer(cols = everything(), names_to = "metric", values_to = "value")

      d_dag_tbl <- tibble::tibble(
        metric = names(d_dag_i),
        d_dag = as.numeric(d_dag_i)
      )

      rand_i_long %>%
        dplyr::group_by(metric) %>%
        dplyr::summarise(
          q025 = quantile(value, probs = 0.025, na.rm = TRUE),
          q975 = quantile(value, probs = 0.975, na.rm = TRUE),
          null_mean = mean(value, na.rm = TRUE),
          .groups = "drop"
        ) %>%
        dplyr::left_join(d_dag_tbl, by = "metric") %>%
        dplyr::mutate(
          algorithm = algorithm,
          gt_id = gt_id,
          in_95_null = d_dag >= q025 & d_dag <= q975
        )
    }, error = function(e) {
      message(
        sprintf(
          "algorithm %s, gt_id %d, attempt %d failed (%s). Regenerating ground-truth DAG.",
          algorithm, gt_id, attempt, conditionMessage(e)
        )
      )
      NULL
    })

    if (!is.null(result_i)) {
      return(result_i)
    }

    attempt <- attempt + 1L
  }
}

set.seed(1)

pc_pcalg <- causalDisco::pc(engine = "pcalg", test = "fisher_z", u2pd = "rand", alpha = 0.05, conservative = FALSE)
ges_pcalg <- causalDisco::ges(engine = "pcalg", score = "sem_bic")
boss_tetrad <- causalDisco::boss(engine = "tetrad", score = "sem_bic", output_cpdag = TRUE)

method_specs <- list(
  list(name = "pc", method = pc_pcalg),
  list(name = "ges", method = ges_pcalg),
  list(name = "boss", method = boss_tetrad)
)
```


```{r}
# Repeat the same workflow over many simulated ground-truth DAGs (all with p = 10)
n_gt <- 15
p_gt <- 10
n_obs <- 10000
e_prob_gt <- 0.3
n_null <- 20
e_prob_null <- 0.3
max_attempts <- 30

multi_gt_eval <- map_dfr(method_specs, function(spec) {
  map_dfr(
    seq_len(n_gt),
    ~ evaluate_one_ground_truth(gt_id = .x, method = spec$method, algorithm = spec$name)
  )
})

```


# Results

Now we discuss results yeah!

```{r}
# Across all 20 ground truths: how often the estimated-DAG metric falls inside the null 95% interval
multi_gt_summary <- multi_gt_eval %>%
  dplyr::group_by(algorithm, metric) %>%
  dplyr::summarise(
    prop_inside_95 = mean(in_95_null),
    mean_d_dag = mean(d_dag),
    mean_null_q025 = mean(q025),
    mean_null_q975 = mean(q975),
    .groups = "drop"
  ) %>%
  dplyr::arrange(metric)

multi_gt_summary
```

```{r}
# Repeat the same workflow over many simulated ground-truth DAGs (all with p = 10)
e_prob_gt <- 0.7
e_prob_null <- 0.7
 
multi_gt_eval_2 <- map_dfr(method_specs, function(spec) {
  map_dfr(
    seq_len(n_gt),
    ~ evaluate_one_ground_truth(gt_id = .x, method = spec$method, algorithm = spec$name)
  )
})
```

Let's see what happens when the random DAG is more dense than the ground truth.

The performance decreased across the board. This is expected as the sparsity parameter/penalty was not adjusted. Furthermore, the relative performance changed, with BOSS performing better than PC and FGES. The purpose is to demonstrate the importance of carefully choosing the sparsity parameter and/or choose an algorithm that performs well across different degrees of sparsity. 
```{r}
multi_gt_summary_2 <- multi_gt_eval_2 %>%
  dplyr::group_by(algorithm, metric) %>%
  dplyr::summarise(
    prop_inside_95 = mean(in_95_null),
    mean_d_dag = mean(d_dag),
    mean_null_q025 = mean(q025),
    mean_null_q975 = mean(q975),
    .groups = "drop"
  ) %>%
  dplyr::arrange(metric)

multi_gt_summary_2
multi_gt_summary

multi_gt_summary[3:6] - multi_gt_summary_2[3:6]

multi_gt

```
## Even more realism
The last thing I want to do is to choose a random exposure and outcome and evaluate the bias of the estimated total effect of the exposure on the outcome using the estimated graph. This is more aligned with the motivating context, where the main interest is in the effects of a limited number of exposures on a limited number of outcomes. 

```{r}
# For each ground truth, randomly select an exposure and outcome and calculate bias of estimated total effect

# Generate ground truth graph

# Estimate graph using causal discovery algorithm

# Choose random exposure and outcome

# Calculate ATE using true graph and estimated graph

# Report bias and confidence interval of bias
```



# Summary
- Motivated the field of causal discovery
- Discussed some of the metrics commonly used to evaluate causal discovery methods
- Evaluated several causal discovery algorithms in a relatively simple setting
- Compared the performance to randomly generating a causal graph

# Conclusions
Now we return to the original question: can causal discovery algorithms be useful for practical problems? The above would suggest that, as a "data-driven" method to "automate" variable selection in applied problems, the answer is no! But in the real-world, one can encode background knowledge to forbid implausible, or enforce known, edges, which may help.

Although, it should again be reiterated that this is in the simplest case using relatively large sample sizes, no missing data, methods that are "correct" for the data type, etc. Real-world data, especially in the health and biomedical sciences is considerably more complex: data includes mixed types, multiple measurements for the same individual, people drop-out from studies and sample sizes can be small. There do not currently exist causal discovery methods that account for all of these complexities. It is therefore an even bigger unknown how useful causal discovery methods can be in these settings. Luckily, this is an active area of research, and the future is promising! 


# Additional notes
- All causal graphs were made using the 'caugi' R package (https://caugi.org/)
- I am using R version 4.5.1





